---
title: "fl_juvenile_markdown"
author: "Riin Aljas"
date: "10/10/2019"
output:
  html_document: default
  pdf_document: default
---
<!-- Timestamp: leave this in to automatically track when a file is rendered -->
*Rendered at `r format(Sys.time(), '%I:%M %p, %b. %d, %Y')`*

<!-- Contact info: leave author/reporter/editor contact info here -->
*Riin Aljas (aljasriin@gmail.com)*
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, 
  message = FALSE, 
  warning = FALSE
  )
knitr::opts_knit$set(
  root.dir = rprojroot::find_rstudio_root_file()
  )
```

```{r data & lbraries}
source("analysis/etl_florida.R")



library(DT)

by_dc <- lifers %>%
  group_by(dc_number, last_name, first_name, middle_name, dob, gender, race) %>%
  count()

```


##**OVERWIEW**

write stuff here 



##**FINDINGS**

 - **What is the overall number of juveniles in each state currently serving a life sentence?**
 
We don't know yet, the numbers of people who have convicted with a life sentence since 1966 is **```r n_distinct(lifers$dc_number)```**

WHAT ABOUT THE PEOPLE WHO HAVE BEEN RELEASED? ARE THEY'RE IN THE DATA OR NOT? (E.G. HERBERT JOHSNON, IS HE IN THE DATA 
BECAUSE OUR DATA IS FROM APRIL OR ACTUALLY THEY'RE THERE GO BACK AND CHECK THAT )

WE NEED TO DO ANALYSIS FOR CURRENT ONES SEPARATELY (MAYBE JOIN WITH RELEASED, WE SHOULD HAVE IT?)

 
 - ** What is the racial breakdown of juvenile lifers in each state?**
```{r }
 datatable(by_dc %>%
  group_by(race) %>%
  summarise(people = n()) %>%
  mutate(pct = people/nrow(by_dc)*100))
```
Additionally, 2.4% of them are women, and 97.5% are male.  50% of all juvenile offenders were 17, 33% 16, 14% 15, 3% 14 and 0.8 13.

 - **Are the racial splits out of proportion to the prison population as a whole, or the state population as a whole?**
 
 
 - ** Has there been a change since 2016, the date of a key supreme court case,  in the number of juvenile lifers in each state?  - ** If so, what is the change. 
 - ** If there has been a change, has the racial distribution of people affected the same as the racial distribution of the juvenile lifer population as a whole? 
 - ** As in, are people of one race being let out at a higher rate than another race? 
 ** If there has been a change, is there a geographic disparity in who is being affected?  
 ** - Are people from one part of a state more likely to get out?
